---
title: "Multilevel Correlations"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, correlation, types]
vignette: >
  %\VignetteIndexEntry{Multilevel Correlations}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---



```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '',
        digits=2,
        dpi=450)


if (!requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
}
```



## Data


Imagine we have an experiment in which **10 individuals** completed a task with **100 trials**. For each of the 1000 total trials, we measured two things, **V1** and **V2**, and our research aims at **investingating the link between these two variables**.

We will generate data using the [`simulate_simpson()`](https://easystats.github.io/correlation/reference/simulate_simpson.html) function from the `correlation` package.


```{r message=FALSE, warning=FALSE}
library(correlation)

data <- simulate_simpson(n=100, groups=10)
```

Now let's visualize the two variables:


```{r message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(data, aes(x=V1, y=V2)) + 
  geom_point() +
  geom_smooth(colour="black", method="lm", se=FALSE) +
  theme_classic()
```

That seems pretty straightfoward! It seems like there is a **negative correlation** between V1 and V2. Let's test this.

## Simple correlation

```{r message=FALSE, warning=FALSE}
correlation(data)
```

Indeed, there is **strong, negative and significant correlation** between V1 and V2. Great, can we go ahead and **publish these results in PNAS**?

## The Simpson's Paradox

Hold on sunshine! Ever heard of something called the [**Simpson's Paradox**](https://en.wikipedia.org/wiki/Simpson%27s_paradox)?

Let's colour our datapoints by group (by individuals):

```{r message=FALSE, warning=FALSE}
library(ggplot2)

ggplot(data, aes(x=V1, y=V2)) + 
  geom_point(aes(colour=Group)) +
  geom_smooth(aes(colour=Group), method="lm", se=FALSE) + 
  geom_smooth(colour="black", method="lm", se=FALSE) + 
  theme_classic()
```

Mmh, interesting. It seems like, for each subject, the relationship is different. The negative general trend seems to be created by **differences between the groups** and could be spurious! 

**Multilevel *(as in multi-group)* correlations allow us to account for differences between groups**. It is based on a partialization of the group, entered as a random effect in a mixed linear regression.

You can compute them with the [**correlations**](https://github.com/easystats/correlation) package by setting the `multilevel` arguent to `TRUE`.


```{r message=FALSE, warning=FALSE}
correlation(data, multilevel = TRUE)
```

**Dayum!** We were too hasty in our conclusions! Taking the group into account seems to be super important.


Note: In this simple case where only two variables are of interest, it would be of course best to directly proceed using a mixed regression model instead of correlations. That being said, the latter can be useful for exploratory analysis, when multiple variables are of interest, or in combination with a network or structural approach.
