---
title: "Correlation Types"
output: 
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, correlation, types]
vignette: >
  %\VignetteIndexEntry{Correlation Types}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r, include=FALSE}
library(knitr)
options(knitr.kable.NA = "")
knitr::opts_chunk$set(
  comment = ">",
  out.width = "100%",
  message = FALSE,
  warning = FALSE,
  dpi = 450
)
options(digits = 2)

set.seed(333)


if (!requireNamespace("see", quietly = TRUE) ||
  !requireNamespace("datawizard", quietly = TRUE) ||
  !requireNamespace("poorman", quietly = TRUE) ||
  !requireNamespace("ggplot2", quietly = TRUE)) {
  knitr::opts_chunk$set(eval = FALSE)
} else {
  library(see)
  library(datawizard)
  library(poorman)
  library(ggplot2)
}
```

---

This vignette can be cited as:

```{r cite}
citation("correlation")
```

---

## Different Methods for Correlations

Correlations tests are arguably one of the most commonly used statistical
procedures, and are used as a basis in many applications such as exploratory
data analysis, structural modeling, data engineering, etc. In this context, we
present **correlation**, a toolbox for the R language [@Rteam] and part of the
[**easystats**](https://github.com/easystats/easystats) collection, focused on
correlation analysis. Its goal is to be lightweight, easy to use, and allows for
the computation of many different kinds of correlations, such as:

- **Pearson's correlation**: This is the most common correlation method. It
  corresponds to the covariance of the two variables normalized (i.e., divided)
  by the product of their standard deviations.

$$r_{xy} = \frac{cov(x,y)}{SD_x \times SD_y}$$

- **Spearman's rank correlation**: A non-parametric measure of correlation, the
Spearman correlation between two variables is equal to the Pearson correlation
between the rank scores of those two variables; while Pearson's correlation
assesses linear relationships, Spearman's correlation assesses monotonic
relationships (whether linear or not). Confidence Intervals (CI) for Spearman's
correlations are computed using the @fieller1957tests correction [see
@bishara2017confidence].

$$r_{s_{xy}} = \frac{cov(rank_x, rank_y)}{SD(rank_x) \times SD(rank_y)}$$

- **Kendall's rank correlation**: In the normal case, the Kendall correlation is
preferred to the Spearman correlation because of a smaller gross error
sensitivity (GES) and a smaller asymptotic variance (AV), making it more robust
and more efficient. However, the interpretation of Kendall's tau is less direct
compared to that of the Spearman's rho, in the sense that it quantifies the
difference between the % of concordant and discordant pairs among all possible
pairwise events. Confidence Intervals (CI) for Kendall's correlations are
computed using the @fieller1957tests correction [see
@bishara2017confidence]. For each pair of observations (i ,j) of two variables
(x, y), it is defined as follows:

$$\tau_{xy} = \frac{2}{n(n-1)}\sum_{i<j}^{}sign(x_i - x_j) \times sign(y_i - y_j)$$

- **Biweight midcorrelation**: A measure of similarity that is median-based,
instead of the traditional mean-based, thus being less sensitive to outliers. It
can be used as a robust alternative to other similarity metrics, such as Pearson
correlation [@langfelder2012fast].

- **Distance correlation**: Distance correlation measures both linear and
non-linear association between two random variables or random vectors (for more, see @szekely2007measuring, @szekely2009brownian). This is
in contrast to Pearson's correlation, which can only detect linear association
between two random variables.

- **Percentage bend correlation**: Introduced by Wilcox (1994), it is based on a
down-weight of a specified percentage of marginal observations deviating from
the median (by default, 20 percent).

- **Shepherd's Pi correlation**: Equivalent to a Spearman's rank correlation
after outliers removal (by means of bootstrapped Mahalanobis distance).

- **Blomqvist’s coefficient**: The Blomqvist’s coefficient (also referred to as
Blomqvist's Beta or medial correlation; Blomqvist, 1950) is a median-based
non-parametric correlation that has some advantages over measures such as
Spearman's or Kendall's estimates (see Shmid and Schimdt, 2006).

- **Hoeffding’s D**: The Hoeffding’s D statistic is a non-parametric rank based
measure of association that detects more general departures from independence
(Hoeffding 1948), including non-linear associations. Hoeffding’s D varies
between -0.5 and 1 (if there are no tied ranks, otherwise it can have lower
values), with larger values indicating a stronger relationship between the
variables.

- **Gamma correlation**: The Goodman-Kruskal gamma statistic is similar to
Kendall's Tau coefficient. It is relatively robust to outliers and deals well
with data that have many ties.

- **Gaussian rank correlation**: The Gaussian rank correlation estimator is a
simple and well-performing alternative for robust rank correlations (Boudt et
al., 2012). It is based on the Gaussian quantiles of the ranks.

- **Point-Biserial and biserial correlation**: Correlation coefficient used when
one variable is continuous and the other is dichotomous (binary). Point-Biserial
is equivalent to a Pearson's correlation, while Biserial should be used when the
binary variable is assumed to have an underlying continuity. For example,
anxiety level can be measured on a continuous scale, but can be classified
dichotomously as high/low.

- **Winsorized correlation**: Correlation of variables that have been
Winsorized, i.e., transformed by limiting extreme values to reduce the effect of
possibly spurious outliers.

- **Polychoric correlation**: Correlation between two theorised normally
distributed continuous latent variables, from two observed ordinal variables.

- **Tetrachoric correlation**: Special case of the polychoric correlation
applicable when both observed variables are dichotomous.

- **Partial correlation**: Correlation between two variables after adjusting for
the (linear) effect of one or more variables. The correlation test is 
run after having partialized the dataset, independently from it. In other words,
it considers partialization as an independent step generating a different
dataset, rather than belonging to the same model. This is why some discrepancies
are to be expected for the *t*- and the *p*-values (but not the correlation
coefficient) compared to other implementations such as `ppcor`. Let $e_{x.z}$ be
the residuals from the linear prediction of $x$ by $z$ (note that this can be
expanded to a multivariate $z$):

$$r_{xy.z} = r_{e_{x.z},e_{y.z}}$$

- **Multilevel correlation**: Multilevel correlations are a special case of
partial correlations where the variable to be adjusted for is a factor and is
included as a random effect in a mixed-effects model.

## Comparison

We will fit different types of correlations of generated data with different
link strengths and link types.

Let's first load the required libraries for this analysis.

```{r}
library(correlation)
library(bayestestR)
library(see)
library(ggplot2)
library(datawizard)
library(poorman)
```

### Utility functions

```{r}
generate_results <- function(r, n = 100, transformation = "none") {
  data <- bayestestR::simulate_correlation(round(n), r = r)

  if (transformation != "none") {
    var <- ifelse(grepl("(", transformation, fixed = TRUE), "data$V2)", "data$V2")
    transformation <- paste0(transformation, var)
    data$V2 <- eval(parse(text = transformation))
  }

  out <- data.frame(n = n, transformation = transformation, r = r)

  out$Pearson <- cor_test(data, "V1", "V2", method = "pearson")$r
  out$Spearman <- cor_test(data, "V1", "V2", method = "spearman")$rho
  out$Kendall <- cor_test(data, "V1", "V2", method = "kendall")$tau
  out$Biweight <- cor_test(data, "V1", "V2", method = "biweight")$r
  out$Distance <- cor_test(data, "V1", "V2", method = "distance")$r
  out$Distance <- cor_test(data, "V1", "V2", method = "distance")$r

  out
}
```

### Effect of Relationship Type

```{r}
data <- data.frame()
for (r in seq(0, 0.999, length.out = 200)) {
  for (n in c(100)) {
    for (transformation in c(
      "none",
      "exp(",
      "log10(1+max(abs(data$V2))+",
      "1/",
      "tan(",
      "sin(",
      "cos(",
      "cos(2*",
      "abs(",
      "data$V2*",
      "data$V2*data$V2*",
      "ifelse(data$V2>0, 1, 0)*("
    )) {
      data <- rbind(data, generate_results(r, n, transformation = transformation))
    }
  }
}

data %>%
  datawizard::reshape_longer(
    select = -c("n", "r", "transformation"),
    names_to = "Type",
    values_to = "Estimation"
  ) %>%
  mutate(Type = relevel(as.factor(Type), "Pearson", "Spearman", "Kendall", "Biweight", "Distance")) %>%
  ggplot(aes(x = r, y = Estimation, fill = Type)) +
  geom_smooth(aes(color = Type), method = "loess", alpha = 0) +
  geom_vline(aes(xintercept = 0.5), linetype = "dashed") +
  geom_hline(aes(yintercept = 0.5), linetype = "dashed") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  see::theme_modern() +
  scale_color_flat_d(palette = "rainbow") +
  scale_fill_flat_d(palette = "rainbow") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  facet_wrap(~transformation)

model <- data %>%
  datawizard::reshape_longer(
    select = -c("n", "r", "transformation"),
    names_to = "Type",
    values_to = "Estimation"
  ) %>%
  lm(r ~ Type / Estimation, data = .) %>%
  parameters::parameters()

arrange(model[6:10, ], desc(Coefficient))
```

As we can see, **distance** correlation is able to capture the strength even for
severely non-linear relationships.

# References

