---
title: "Correlation Types"
output: 
  github_document:
    toc: true
    fig_width: 10.08
    fig_height: 6
  rmarkdown::html_vignette:
    toc: true
    fig_width: 10.08
    fig_height: 6
tags: [r, correlation, types]
vignette: >
  %\VignetteIndexEntry{Correlation Types}
  \usepackage[utf8]{inputenc}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: bibliography.bib
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(knitr)
options(knitr.kable.NA = '')
knitr::opts_chunk$set(comment=">",
  dpi=450)
options(digits=2)

set.seed(333)
```




## Different Methods for Correlations

<!-- This is copied from the details section of the documentation of the correlation function, any changes here or there should be SYNCED -->

- **Pearson's correlation**: The covariance of the two variables divided by the product of their standard deviations.

- **Spearman's rank correlation**: A nonparametric measure of rank correlation (statistical dependence between the rankings of two variables). The Spearman correlation between two variables is equal to the Pearson correlation between the rank values of those two variables; while Pearson's correlation assesses linear relationships, Spearman's correlation assesses monotonic relationships (whether linear or not).

- **Kendall's rank correlation**: In the normal case, the Kendall correlation is preferred than the Spearman correlation because of a smaller gross error sensitivity (GES) and a smaller asymptotic variance (AV), making it more robust and more efficient. However, the interpretation of Kendall's tau is less direct than that of Spearman's rho, in the sense that it quantifies the difference between the % of concordant and discordant pairs among all possible pairwise events.

- **Biweight midcorrelation**: A measure of similarity between samples that is median-based, rather than mean-based, thus is less sensitive to outliers, and can be a robust alternative to other similarity metrics, such as Pearson correlation.

- **Distance correlation**: Distance correlation measures both linear and nonlinear association between two random variables or random vectors. This is in contrast to Pearson's correlation, which can only detect linear association between two random variables.

- **Percentage bend correlation**: Introduced by Wilcox (1994), it is based on a down-weight of a specified percentage of marginal observations deviating from the median (by default, 20%).

- **Shepherd's Pi correlation**: Equivalent to a Spearman's rank correlation after outliers removal (by means of bootstrapped mahalanobis distance).

- **Point-Biserial and biserial correlation**: Correlation coefficient used when one variable is continuous and the other is dichotomous (binary). Point-serial is equivalent to a Pearson's correlation, while Biserial should be used when the binary variable is assumed to have an underlying continuity. For example, anxiety level can be measured on a continuous scale, but can be classified dichotomously as high/low.

- **Polychoric correlation**: Correlation between two theorised normally distributed continuous latent variables, from two observed ordinal variables.

- **Tetrachoric correlation**: Special case of the polychoric correlation applicable when both observed variables are dichotomous.




## Comparison


```{r message=FALSE, warning=FALSE}
library(bayestestR)
library(correlation)
library(ggplot2)
library(tidyr)
library(see)
```

We will fit different types of correlations of generated data with different link strengths and link types.

### Utility functions

```{r message=FALSE, warning=FALSE}
generate_results <- function(r, n = 100, transformation = "none"){
  data <- bayestestR::simulate_correlation(round(n), r = r)
  if(transformation != "none"){
    var <- ifelse(grepl("(", transformation, fixed = TRUE), "data$V2)", "data$V2")
    transformation <- paste0(transformation, var)
    data$V2 <- eval(parse(text = transformation))
  }
  out <- data.frame(n = n, transformation = transformation, r = r)
  
  out$Pearson <- cor_test(data, "V1", "V2", method = "pearson")$r
  out$Spearman <- cor_test(data, "V1", "V2", method = "spearman")$rho
  out$Kendall <- cor_test(data, "V1", "V2", method = "kendall")$tau
  out$Biweight <- cor_test(data, "V1", "V2", method = "biweight")$r
  out$Distance <- cor_test(data, "V1", "V2", method = "distance")$r
  out$Distance <- cor_test(data, "V1", "V2", method = "distance")$r
  out
}
```


### Effect of Relationship Type

```{r message=FALSE, warning=FALSE}
data <- data.frame()
for(r in seq(0, 0.999, length.out = 200)){
  for(n in c(100)){
    for(transformation in c("none", "exp(", "log10(1+max(abs(data$V2))+", "1/", "tan(", "sin(", "cos(", "cos(2*","abs(", "data$V2*", "data$V2*data$V2*", "ifelse(data$V2>0, 1, 0)*(")){
      data <- rbind(data, generate_results(r, n, transformation = transformation))
    }
  }
}


data %>% 
  tidyr::pivot_longer(-c(n, r, transformation), names_to = "Type", values_to = "Estimation") %>%
  dplyr::mutate(Type = forcats::fct_relevel(Type, "Pearson", "Spearman", "Kendall", "Biweight", "Distance")) %>% 
  ggplot(aes(x = r, y = Estimation, fill = Type)) +
  geom_smooth(aes(color = Type), method = 'loess', alpha = 0) +
  geom_vline(aes(xintercept = 0.5), linetype = "dashed") +
  geom_hline(aes(yintercept = 0.5), linetype = "dashed") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  see::theme_modern() +
  scale_color_flat_d(palette = "rainbow") +
  scale_fill_flat_d(palette = "rainbow") +
  guides(colour = guide_legend(override.aes = list(alpha = 1))) +
  facet_wrap(~transformation)

model <- data %>% 
  tidyr::pivot_longer(-c(n, r, transformation), names_to = "Type", values_to = "Estimation") %>%
  lm(r ~ Type / Estimation, data = .) %>% 
  parameters::parameters()
dplyr::arrange(model[6:10,], desc(Coefficient))
```

As we can see, **distance** correlation is able to capture the strength even for severly non-linear relationships.


# References
